Continous time Markov chain models; \\ extension, simulation and visualisation

### 2015-02-11 Response.tex

\begin{quote}
\myeditsvthree
\end{quote}

% 4
\item {
{\it
\textbf{
Expression (1.12) is confusing in that indices i and j appear in the lefthand side, but not in the righthand side. 
Likewise in expression (1.13).
}% END: textbf
}% END: it
}% END: item
\\
Corrected.

% 5
\item {
{\it
\textbf{
P20, 2nd display: an equality sign is missing on the 2nd line.
}% END: textbf
}% END: it
}% END: item
\\
Corrected.

% 6
\item {
{\it
\textbf{
I would find it useful to see a brief discussion on the plausibility of (1.22).
}% END: textbf
}% END: it
}% END: item
\\
% FB: Does he mean plausibility from a biological perspective or statistical? It's a necessary condition for the existence of stationary dist.
% Pl: I see it as questioning the plausibility in a biological sense. We now already argue that it is an assumption taken for computational reasons, but at the end of that paragraph we can explicitly state 'Realistic substitution processes are however unlikely to be time-reversible and procedures have also been developed to compute phylogenetic likelihoods under non-reversible models (cite Boussau and Gouy, 2006, see below)'. ...perhaps add in something more from that paper.... and end with 'We note that we use such non-reversible models in our discrete phytogeographic applications (cite Edwards et al., Current Biology, 2011).
% http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.324.620&rep=rep1&type=pdf
Please see my changes which explain how the assumptions of stationarity and homogeneity of stochastic process governing the evolution connect to each other  
%PL: Please fix here
with the assumption of

and which start with:

@@

% 7
\item {
{\it
\textbf{
If I am right that matrix P(t) on the bottom of page 22 is four-dimensional, then why not write it in full, which will be less confusing?
}% END: textbf
}% END: it
}% END: item
\\
Corrected.

% 8
\item {
{\it
\textbf{
P25, the statement at the end of the penultimate paragraph (`most plausible hypothesis for the data') is vague.
}% END: textbf
}% END: it
}% END: item
\\
This statement now reads:

\begin{quote}
\myeditsveight
\end{quote}

% 9
\item {
{\it
\textbf{
P28, I understand what T(n) is, but would nevertheless find it helpful if it were stated more explicitly.
}% END: textbf
}% END: it
}% END: item
\\
% FB: I don't understand the question in the definition we clearly define T as being simply a function on a positive subset of real numbers?
% Does he mean we should wrote "time"? but this is not always true?
% PL: OK, but can we then state something like this: T(n) can be interpreted as time in most practical situations, but this does necessarily need to be the case, e.g.....?
We now clarified that T(n) represents the ``worst-case' time' complexity of an algorithm.

@@

% 10
\item {
{\it
\textbf{
P31, If the phrase `proper prior' is used, then perhaps it should be defined.
}% END: textbf
}% END: it
}% END: item
\\
This paragraph now reads:

\begin{quote}
\myeditsvnine
\end{quote}

% 11
\item {
{\it
\textbf{
P32, Since the author assumes that the reader may have no prior knowledge of Bayesian inference, I believe the meaning of Figure 1.11 should be explained in more detail.
}% END: textbf
}% END: it
}% END: item
\\
Corrected. There is now a more elaborate description accompanying this figure (see Subsection~\ref{sub:ex_beta_binom}).

@@

% 12
\item {
{\it
\textbf{
P33,. I think the notation $\theta[t]$ has not been defined.
}% END: textbf
}% END: it
}% END: item
\\
Corrected.

% 13
\item {
{\it
\textbf{
There is a typo in the lefthand side of (1.37).
}% END: textbf
}% END: it
}% END: item
\\
Corrected.

% 14
\item {
{\it
\textbf{
I would find it useful to get a bit more detail how a model for the likelihood of Y and for the prior of $\Phi$ might look like.
}% END: textbf
}% END: it
}% END: item
\\
% FB: I'm not sure if I understand this comment, we do write that the likelihood is standard tree pruning and there's a whole subsection in the Introduction that delves on that...  We could elaborate on the prior choice here  ...
% PL: OK, I would simply point out that we use exactly the same type of CTMC modelling in the response letter. And simply to do something in response to the comment, modify this sentence: "Unlike before, the rate matrix does not necessarily have to be symmetrical, i.e. there exist some i, j such that ..." into "So, we employ the same type of CTMC modelling for discrete phylogeography as for nucleotide substitution processes, which is described in detail in ...(refer to intro section on CTMC), but now for an arbitrary number of location states. And unlike for nucleotide substitution, the rate matrix does not necessarily have to be symmetrical, i.e. there exist some i, j such that ..." 
In response to a comment above, we now provide more detail in Subsection~\ref{sub:phylogeo}.

% 15
\item {
{\it
\textbf{
P43. What is compositional heterogeneity?
}% END: textbf
}% END: it
}% END: item
\\
We now explicitly state that this refers to different stationary frequencies for the CTMC.

@@

\begin{quote}
\myeditsvfifteen
\end{quote}

% 16
\item {
{\it
\textbf{
P44. I would find it useful to receive a bit more guidance concerning the connection to phylogeographic inference. 
Also in Section 2.4.3, the description of the proposal by Bahl et al. (2011) is too brief for me to appreciate why the epoch model is more appropriate.
}% END: textbf
}% END: it
}% END: item
\\
% FB: This becomes p46 now, from "The former is critical to accommodate discrete phylogeographic inference"
In addition to the detail added to the phylogeography section in the Introduction, I also added a sentence to clarify the approach taken by Bahl et al.:

@@

\begin{quote}
\myeditsvsixteen
\end{quote}

This further clarifies the distinction between the \emph{ad-hoc} method by Bahl et al. and the ``epoch model'' applied to the global influenza spread problem.

% 17
\item {
{\it
\textbf{
Display (2.5): I believe this display implicitly assumes that the time T1 is a priori known. 
It would be good to state this explicitly.
}% END: textbf
}% END: it
}% END: item
\\
This is now states explicitly.
%PL: where?may be quote here?

% 18
\item {
{\it
\textbf{
P57. The sentence `To determine the posterior odds, ... of that model' is vague.
}% END: textbf
}% END: it
}% END: item
\\
The sentence is rephrased to:

@@

\begin{quote}
\myeditsveighteen
\end{quote}

% 19
\item {
{\it
\textbf{
P85. Riemannian -> Riemann
}% END: textbf
}% END: it
}% END: item
\\
Corrected.

% 20
\item {
{\it
\textbf{
P86. The last sentence of Section 5.1.3 is not sufficiently clear. 
Please make clear why the use of mixed effects models could allow this.
}% END: textbf
}% END: it
}% END: item
\\
I changed Subsection~\ref{sub:glm_epoch}, explaining that by this sentence we mean the impact of various covariates,  and that by using the MG94 parametrization we do not need to assume altho all rate variation is due variation in non-synonymous rates.
%PL: why not quote it (or at least part of it)?
There is also a more elaborate description of the model we plan to use in the upcoming manuscript on the application of the hierarchical epoch model.

% 21
\item {
{\it
\textbf{
P88. It would be useful to receive a bit more detail what is meant by `infinite hidden Markov Model'.
}% END: textbf
}% END: it
}% END: item
\\
I introduce the definition of HMM's in a form of Remark, please see Subsection~\ref{sub:dpp}.

### 2015-02-11 chapters/chapter01.tex
response, dated tip model

\myedit{svoneb}{
Time-stamped or `heterchronous' sequence data, which frequently arises from studies involving MEP populations for which sequences are sampled at different points in time and the evolutionary rates are high enough for significant changes to occur between the sampling times, can be used to calibrate the molecular clock and convert branch lengths into actual time units.
}
For simplicity let us assume a single site, where character sequence data was sampled on dates ranging from $1983$ to $2000$, as seen in Figure~\ref{fig:timing_sequences}.
We also assume that the rooted tree topology is known and we want to estimate the unknown times $T_{0},T_{6},T_{7},T_{8}$.

@@

We start by estimating the rate of mutation $r$, which here is assumed to apply to the whole tree (a strict clock), from the differences in sampling times and the genetic distance, given by the model of sequence substitution.
This allows us to change the scale of the tree from units of substitutions to a time-scale measured in years.
The likelihood of the tree given estimated parameters $r,T_{0},T_{6},T_{7},T_{8}$ can be calculated using methods presented in Subsection~\ref{sub:likelihood}.

>>>>

Time-stamped or `heterchronous' sequence data, which frequently arises from studies involving MEP populations for which sequences are sampled at different points in time and the evolutionary rates are high enough for significant changes to occur between the sampling times, can be used to calibrate the molecular clock and convert branch lengths into actual time units.
For simplicity let us assume a single site, where character sequence data was sampled on dates ranging from $1983$ to $2000$, as seen in Figure~\ref{fig:timing_sequences}.
We also assume that the rooted tree topology is known and we want to estimate the unknown times $T_{0},T_{6},T_{7},T_{8}$.

@@

\myedit{svoneb}{
We start by considering heights of tree tips and introducing a new parameter, the rate of mutation $r$, which here for simplicity is assumed to apply to the whole tree (a strict clock), which allows us to change the scale of the tree from units of substitutions to a time-scale measured in years.

Both the rate of evolution $r$ and the unobserved heights $T_{0},T_{6},T_{7},T_{8}$ are estimated from the molecular sequence data, with the constraint that the heights are proportional to their sampling time and teh ratio of this proportionality being $r$.
}
